A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-14 paper.Rmd
tweaked the intro and placement of fig 1

Diagnostics are the key to determining whether there is anything importantly wrong with a model. In linear regression analysis, residuals from the model fit are commonly used. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. Residuals may be plotted as a histogram or quantile-quantile plot to examine the distribution. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we consider it to be well behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there is a cause for concern. One could also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk (SW) Normality test [@shapiro1965analysis].

More typically, a scatterplot of residuals plotted against the fitted values and each of the explanatory variables, is made to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specifed. In general, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. However, correctly judging a residual plot where no pattern exists can be a difficult task for humans. It is especially common, particularly among students, to misinterpret patterns that are random noise and random deviation from a model [@loy2021bringing]. There are also numerous conventional hypothesis tests available to test for non-linear dependence [@ramsey_tests_1969], and heteroskedasticity  [@breusch_simple_1979].

An abundance of literature describe appropriate diagnostic methods for linear regression, e.g. @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. All these writings consider plotting residuals to be an essential tool for diagnosing regression modelling problems. In addition, @draper1998applied and @belsley_regression_1980 believe that residual plots are usually revealing when the assumptions are violated. @cook_applied_1999 thinks formal tests and graphical procedures are complementary and both have a place in residual analysis, but they focus on graphical methods rather than on formal testing, as they are easier to use. @montgomery1982introduction even suggests that residual plots are more informative in most practical situations than the corresponding conventional hypothesis tests.

A common guidance by experts is that optimal method for diagnosing model fits is by plotting the data. The persistence of this advice to check the plots is curious, and investigating why this might be common advice is the subject of this paper. The paper is structured as follows. The next section describes the background on the types of departures that one expects to detect, and outlines a formal statistical process for reading residual plots, called visual inference. Section \ref{experimental-design} details the experimental design to compare the decisions made by formal hypothesis testing, and how humans would read diagnostic plots. The results are reported in Section \ref{results}. We conclude with a discussion of future work, in particular, how the responsibility for residual plot reading might be relieved.

@@

```{r residual-plot-common-departures, fig.width = 10, fig.height = 2.5, fig.cap = "Example residual vs fitted value plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot."}

>>>>

Diagnostics are the key to determining whether there is anything importantly wrong with a model. In linear regression analysis, studying the residuals from a model fit is a common diagnostic activity. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their univariate distribution, residuals may be plotted as a histogram or normal probability plot. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we would consider it to be well-behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there would be cause for concern. One could also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk (SW) Normality test [@shapiro1965analysis].

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, are made to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. In general, one looks for noticeable departures from the model such as non-linear dependency or heteroskedasticity. A non-linear dependency would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. We humans will almost always will see a pattern, so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To answer this, there have been numerous conventional hypothesis tests made available to test for non-linear dependence [e.g. @ramsey_tests_1969], and heteroskedasticity  [e.g. @breusch_simple_1979].

Linear regression is a well-established procedure, and there is considerable literature describing diagnostic procedures, e.g. @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. An interesting details, is that despite the abundance of conventional tests, ALL of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems. @draper1998applied and @belsley_regression_1980 explain that residual plots are usually revealing when the assumptions are violated. @cook_applied_1999 thinks formal tests and graphical procedures are complementary and both have a place in residual analysis, but they focus on graphical methods rather than on formal testing. @montgomery1982introduction even suggests that residual plots are more informative in most practical situations than the corresponding conventional hypothesis tests.

The common wisdom of experts is that the optimal method for diagnosing model fits is by plotting the data. The persistence of this advice to check the plots is *curious*, and investigating why this is, is the subject of this paper.

The paper is structured as follows. The next section describes the background on the types of departures that one expects to detect, and outlines a formal statistical process for reading residual plots, called visual inference. Section \ref{experimental-design} details the experimental design to compare the decisions made by formal hypothesis testing, and how humans would read diagnostic plots. The results are reported in Section \ref{results}. We conclude with a discussion of future work, in particular, how the responsibility for residual plot reading might be relieved.

@@

```{r residual-plot-common-departures, fig.pos="t!", fig.width = 10, fig.height = 2.5, fig.cap = "Example residual vs fitted value plots (red line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot."}

### 2023-04-14 paper.tex
tweaked the intro and placement of fig 1

Diagnostics are the key to determining whether there is anything
importantly wrong with a model. In linear regression analysis, residuals
from the model fit are commonly used. Residuals summarise what is not
captured by the model, and thus provide the capacity to identify what
might be wrong.

We can assess residuals in multiple ways. Residuals may be plotted as a
histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, we consider it to be well
behaved. However, if the distribution is skewed, bimodal, multimodal, or
contains outliers, there is a cause for concern. One could also inspect
the distribution by conducting a goodness-of-fit test, such as the
Shapiro-Wilk (SW) Normality test \citep{shapiro1965analysis}.

More typically, a scatterplot of residuals plotted against the fitted
values and each of the explanatory variables, is made to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially inadequate or incorrectly specifed. In general,
one looks for noticeable departures from the model like non-linear
dependency or heteroskedasticity. However, correctly judging a residual
plot where no pattern exists can be a difficult task for humans. It is
especially common, particularly among students, to misinterpret patterns
that are random noise and random deviation from a model
\citep{loy2021bringing}. There are also numerous conventional hypothesis
tests available to test for non-linear dependence
\citep{ramsey_tests_1969}, and heteroskedasticity
\citep{breusch_simple_1979}.

An abundance of literature describe appropriate diagnostic methods for
linear regression, e.g. \citet{draper1998applied},
\citet{montgomery1982introduction}, \citet{belsley_regression_1980},
\citet{cook_applied_1999} and \citet{cook1982residuals}. All these
writings consider plotting residuals to be an essential tool for
diagnosing regression modelling problems. In addition,
\citet{draper1998applied} and \citet{belsley_regression_1980} believe
that residual plots are usually revealing when the assumptions are
violated. \citet{cook_applied_1999} thinks formal tests and graphical
procedures are complementary and both have a place in residual analysis,
but they focus on graphical methods rather than on formal testing, as
they are easier to use. \citet{montgomery1982introduction} even suggests
that residual plots are more informative in most practical situations
than the corresponding conventional hypothesis tests.

A common guidance by experts is that optimal method for diagnosing model
fits is by plotting the data. The persistence of this advice to check
the plots is curious, and investigating why this might be common advice
is the subject of this paper. The paper is structured as follows. The
next section describes the background on the types of departures that
one expects to detect, and outlines a formal statistical process for
reading residual plots, called visual inference. Section
\ref{experimental-design} details the experimental design to compare the
decisions made by formal hypothesis testing, and how humans would read
diagnostic plots. The results are reported in Section \ref{results}. We
conclude with a discussion of future work, in particular, how the
responsibility for residual plot reading might be relieved.

@@

\caption{Example residual vs fitted value plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot.}\label{fig:residual-plot-common-departures}
\end{figure}

>>>>

Diagnostics are the key to determining whether there is anything
importantly wrong with a model. In linear regression analysis, studying
the residuals from a model fit is a common diagnostic activity.
Residuals summarise what is not captured by the model, and thus provide
the capacity to identify what might be wrong.

We can assess residuals in multiple ways. To examine the their
univariate distribution, residuals may be plotted as a histogram or
normal probability plot. Using the classical normal linear regression
model as an example, if the distribution is symmetric and unimodal, we
would consider it to be well-behaved. However, if the distribution is
skewed, bimodal, multimodal, or contains outliers, there would be cause
for concern. One could also inspect the distribution by conducting a
goodness-of-fit test, such as the Shapiro-Wilk (SW) Normality test
\citep{shapiro1965analysis}.

In addition, scatterplots of residuals plotted against the fitted values
and each of the explanatory variables, are made to scrutinize their
relationships. If there are any visually discoverable patterns, the
model is potentially inadequate or incorrectly specified. In general,
one looks for noticeable departures from the model such as non-linear
dependency or heteroskedasticity. A non-linear dependency would suggest
that the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption. However,
correctly judging whether NO pattern exists in a residual plot is a
difficult task for humans. We humans will almost always will see a
pattern, so the question that really needs answering is whether any
pattern perceived is consistent with randomness, purely sampling
variability or noise. It is especially difficult to teach this to new
analysts and students \citep{loy2021bringing}. To answer this, there
have been numerous conventional hypothesis tests made available to test
for non-linear dependence \citep[e.g.][]{ramsey_tests_1969}, and
heteroskedasticity \citep[e.g.][]{breusch_simple_1979}.

Linear regression is a well-established procedure, and there is
considerable literature describing diagnostic procedures, e.g.
\citet{draper1998applied}, \citet{montgomery1982introduction},
\citet{belsley_regression_1980}, \citet{cook_applied_1999} and
\citet{cook1982residuals}. An interesting details, is that despite the
abundance of conventional tests, ALL of these writings advise that
plotting residuals is an essential tool for diagnosing regression model
problems. \citet{draper1998applied} and \citet{belsley_regression_1980}
explain that residual plots are usually revealing when the assumptions
are violated. \citet{cook_applied_1999} thinks formal tests and
graphical procedures are complementary and both have a place in residual
analysis, but they focus on graphical methods rather than on formal
testing. \citet{montgomery1982introduction} even suggests that residual
plots are more informative in most practical situations than the
corresponding conventional hypothesis tests.

The common wisdom of experts is that the optimal method for diagnosing
model fits is by plotting the data. The persistence of this advice to
check the plots is \emph{curious}, and investigating why this is, is the
subject of this paper.

The paper is structured as follows. The next section describes the
background on the types of departures that one expects to detect, and
outlines a formal statistical process for reading residual plots, called
visual inference. Section \ref{experimental-design} details the
experimental design to compare the decisions made by formal hypothesis
testing, and how humans would read diagnostic plots. The results are
reported in Section \ref{results}. We conclude with a discussion of
future work, in particular, how the responsibility for residual plot
reading might be relieved.

@@

\caption{Example residual vs fitted value plots (red line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot.}\label{fig:residual-plot-common-departures}
\end{figure}
